Clonal interference between aneuploidy and mutation

### 2021-04-18 ms/ms.tex
added Wright-Fisher steps for single model

\paragraph*{Evolutionary Models.} To examine the role of aneuploidy in adaptation, two models were developed - \emph{single-mutant model} and \emph{multiple-locus model}. 
Both are based on the Wright-Fisher model~\citep{Tenaillon1999} and include the effects of natural selection, genetic drift, aneuploidy, and mutation. Only beneficial mutations assumed, 
thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication).

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}
      \caption{
        \textbf{A. Single-mutant}
        }
  \end{subfigure}%
  ~ 
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}
      \caption{
        \textbf{B. Multiple-locus}
      }
  \end{subfigure}
  \caption{
    \textbf{Model Schemes.} (A) The population with four possible genotypes and four possible transitions from one genotype to another with the transition rates $\mu$, $\alpha$, 
    $\alpha_l$, $k\mu$. \emph{2n}: euploid wildtype – the initial genotype; \emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; 
    \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; \emph{2n+1*}: 
    aneuploid mutant, with both the beneficial mutation and the extra chromosome. Overall there are two possible paths to reach \emph{2n*} from \emph{2n}.
    (B) Each genotype can have zero or more beneficial mutations and zero or one extra chromosomes. $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate
  }
  \label{fig:models}
\end{figure}

@@

The parameters of the model are: 
$N$, the population size, which is constant in all generations; $\mu$, the mutation rate - the transition rate from \emph{2n} to \emph{2n*}, i.e. the frequency at which cells with genotype \emph{2n*}
are created from cells with genotype \emph{2n}; $\alpha$, the aneuploidy gain rate - the transition rate from \emph{2n} to \emph{2n+1}; 
$\alpha_l$, the aneuploidy loss rate - the transition rate from \emph{2n+1*} to \emph{2n*}; $k$, the coefficient, such that $k\cdot\mu$ is the transition rate from \emph{2n+1} to \emph{2n+1*}, 
where on default $k$ is equals to $33/32$ (euploid \emph{2n} genotyped \emph{S. cerevisiae} contains 32 chromosomes, therefore, assuming each chromosome have equal probability to have the specific 
beneficial mutation in, aneuploid \emph{2n+1} genotype should have 33/32 times the mutation rate); $s$, the mutation selection coefficient; $s_b$, fitness benefit of aneuploidy; 
$c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
\begin{equation} \label{eq:model} \begin{aligned}
w_{2n} & = 1 \\
w_{2n+1} & = 1-c+s_b \\
w_{2n+1*} & = (1+s)(1-c)+s_b \\
w_{2n*} & = (1+s) \;
\end{aligned}
\end{equation}

>>>>

\paragraph*{Evolutionary Models.} To examine the role of aneuploidy in adaptation, two models were developed - \emph{single-mutant model} and \emph{multiple-locus model}. 
Both are based on the Wright-Fisher model~\citep{Tenaillon1999} and include the effects of natural selection, genetic drift, aneuploidy, and mutation. Only beneficial mutations assumed, 
thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication). (\autoref{fig:models})

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}
      \caption{
        \textbf{A. Single-mutant}
        }
      \label{fig:model1}
  \end{subfigure}%
  ~ 
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}
      \caption{
        \textbf{B. Multiple-locus}
      }
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model Schemes.} (A) The population with four possible genotypes and four possible transitions from one genotype to another with the transition rates $\mu$, $\alpha$, 
    $\alpha_l$, $k\mu$. \emph{2n}: euploid wildtype – the initial genotype; \emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; 
    \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; \emph{2n+1*}: 
    aneuploid mutant, with both the beneficial mutation and the extra chromosome. Overall there are two possible paths to reach \emph{2n*} from \emph{2n}.
    (B) Each genotype can have zero or more beneficial mutations and zero or one extra chromosomes. $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate
  }
  \label{fig:models}
\end{figure}

@@

The parameters of the model are: 
$N$, the population size, which is constant in all generations; $\mu$, the mutation rate - the transition rate from \emph{2n} to \emph{2n*}, i.e. the frequency at which cells with genotype \emph{2n*}
are created from cells with genotype \emph{2n}; $\alpha$, the aneuploidy gain rate - the transition rate from \emph{2n} to \emph{2n+1}; 
$\alpha_l$, the aneuploidy loss rate - the transition rate from \emph{2n+1*} to \emph{2n*}; $k$, the coefficient, such that $k\cdot\mu$ is the transition rate from \emph{2n+1} to \emph{2n+1*}, 
where on default $k$ is equals to $33/32$ (euploid \emph{2n} genotyped \emph{S. cerevisiae} contains 32 chromosomes, therefore, assuming each chromosome have equal probability to have the specific 
beneficial mutation in, aneuploid \emph{2n+1} genotype should have 33/32 times the mutation rate); $s$, the mutation selection coefficient; $s_b$, fitness benefit of aneuploidy; 
$c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
\begin{equation} \label{eq:single-w} \begin{aligned}
w_{2n} & = 1 \\
w_{2n+1} & = 1-c+s_b \\
w_{2n+1*} & = (1+s)(1-c)+s_b \\
w_{2n*} & = (1+s) \;
\end{aligned}
\end{equation}

\pagebreak

The model follows Wright-Fisher \citep{Tenaillon1999} steps. The first generation is initialized with $N$ cells of the genotype \emph{2n}. At each generation, the frequency $f_g$ of 
each genotype $g$ is updated:
\begin{enumerate}
  \item Selection step: 
    \begin{equation} \label{eq:single-selection} 
      f_g\gets{\dfrac{f_g w_g}{\sum_{j}f_j w_j}}
      %  j\in\{2n, 2n+1, 2n+1*, 2n*\} 
    \end{equation}
  \item Mutation gain and aneuploidy gain/loss step:
    \begin{equation} \label{eq:single-mutation} \begin{aligned}
      f_{2n} & \gets{f_{2n} - \alpha f_{2n} - \mu f_{2n}}  \\
      f_{2n+1} & \gets{f_{2n+1} - k\mu f_{2n+1} + \alpha f_{2n}}  \\
      f_{2n+1*} & \gets{f_{2n+1*} - \alpha_l f_{2n+1*} + k\mu f_{2n+1}}  \\
      f_{2n*} & \gets{\mu f_{2n} + \alpha_l f_{2n+1}}  \;
    \end{aligned}
    \end{equation}
  \item Random drift step: draw the frequencies of all genotypes from a multinomial distribution
    \begin{equation} \label{eq:single-mutation} 
      f \sim Mult(N, (f_{2n},f_{2n+1},f_{2n+1*},f_{2n*}))
    \end{equation}
\end{enumerate}

\paragraph*{Multiple-locus model.} It expands the \emph{single-mutant model} by allowing for any number of beneficial mutations.
